---
title: "-"
output:
  html_document
---


# Homogeneous Impact Areas

<br />

_Author and date:_
Anders Kolstad
```{r, echo=FALSE}
Sys.Date()
```

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(tmap)
library(dplyr)
#library(osmplotr)
```



## Introduction
This chapter documents the creation of a wall-to-wall map of homogeneous impact areas in Norway.
The map is produces by binning values in the infrastructure index into four discrete categories. 
It is likely to be a good predictor for some indicators, such as  *slitasje* and the presence of
*alien species*.

## About the underlying data
The infrastructure index is explained [here](https://brage.nina.no/nina-xmlui/handle/11250/2983607). 
It is a wall-to-wall raster over Norway with 100 m resolution. Each pixel is given a value along a continuous gradient 
from 0 to around 15.4, representing the frequency or surrounding cells within 500 m with human infrastructure (houses, roads, ect).

### Representativity in time and space
The infrastructure index is calculated with data that have slightly different dates, but can be said to represent the current status, the period 2010-2020.
The data covers all of mainland Norway.


## References
Bakkestuen, V., Dervo, B.K., Bærum, K.M. og Erikstad, L. 2022. Prediksjonsmodellering av naturtyper i ferskvann. NINA Rapport 2079. Norsk institutt for naturforskning.


## Analyses
### Import data

The path must be conditional
```{r}
dir <- substr(getwd(), 1,2)
```



```{r}
path <- ifelse(dir == "C:", 
        "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
        "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif")
```

Import a stars proxy (no data imported yet)
```{r}
infra <- stars::read_stars(path)
```

```{r}
st_crs(infra)$input
```


### Trondheim example
It's easier to see what's happening if we zoom in a bit. Lets get a
boundary box around Trondheim.
```{r}
myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
                ymin = 7032142.5, ymax = 7045245.27),
                crs = st_crs(infra))
```

Cropping the raster to the bbox
```{r}
infra_trd <- sf::st_crop(infra, myBB)
```

Transform to lat long due to osm requirements
```{r, warning=F}
infra_trd_ll <- sf::st_transform(infra_trd, 4326)
```

Get the boundary box of the cropped raster
```{r}
myBB_ll <- sf::st_bbox(infra_trd_ll)
```

### Get OSM highways
Lets download some base maps to help validate and contextualize the infrastructure index.

Download highways using the above bbox.
```{r, warning=FALSE, message=F, eval=F}
hw <- 
  osmplotr::extract_osm_objects(
    bbox = myBB_ll,
    key = "highway",
    sf = T)
```

Transforming highway data back into utm, although not strictly
necessary.
```{r, eval=F}
hw_utm <- sf::st_transform(hw, sf::st_crs(infra_trd)) 
```

This object contains too many roads (about 30k). I'll take out the unnamed roads.
```{r, eval=F}
hw_utm <- hw_utm[!is.na(hw_utm$name),]
```

```{r, eval=F, echo=F}
# Caching the sf object
saveRDS(hw_utm, "data/cache/highways_trondheim.rds")
```

```{r, echo=F}
#Load object
hw_utm <- readRDS("data/cache/highways_trondheim.rds")
```


### Discretize
Here I define a simplified categorical typology for the infrastructure index using four classes. 

```{r}
names(infra_trd) <- "infrastructureIndex"
infra_trd_reclassed <-  infra_trd %>%
  mutate(infrastructureIndex = case_when(
    infrastructureIndex < 1 ~ 0,
    infrastructureIndex < 6 ~ 1,
    infrastructureIndex < 9 ~ 2,
    infrastructureIndex >= 9 ~ 3
  ))
```

Lets plot these two maps side by side
```{r}
map_trd_reclassed <- tm_shape(infra_trd_reclassed)+
  tm_raster(title="Infrastructure Index",
            #palette = "viridis",
            style="cat")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="white")
```

```{r}
map_trd <- tm_shape(infra_trd)+
  tm_raster(title="Infrastructure Index",
            style="cont",
            palette = "viridis")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="white")
```

```{r, fig.cap="Infrastrcture indeks over Trondheim, comparing the continous scale with the ordinal four-step scale. Major roads are in white."}
tmap_arrange(map_trd,
             map_trd_reclassed,
             ncol=1)
```

I tweaked the thresholds for the bins so that most of the forest next to Trondheim (to the left in the map) was in the second lowest class. 
This looks pretty good to me. It depicts a gradient in human presence from high within the build zone, to _no to very low_ in the forests and mountains 
to the west. Note that there is still considerable human activity also there in the form of outdoor recreation and even forestry.

## Discretize the entire map
```{r}
names(infra) <- "infrastructureIndex"

infra_discrete <- infra %>%
  mutate(infrastructureIndex = case_when(
    infrastructureIndex < 1 ~ 0,
    infrastructureIndex < 6 ~ 1,
    infrastructureIndex < 9 ~ 2,
    infrastructureIndex >= 9 ~ 3
  ))
```


## Aggregate
The resolution in this map is more than we need, and the size of the data implies that the next step, the vectorization, would take too long.
We therefore aggregate, ar reduce the resolution.


## Prepare export
<!-- Text here -->


### Eksport file (final product)
<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->




