# Scaling functions

```{r, include=F}
library(ggplot2)
library(ggridges)
library(data.table)
theme_set(theme_bw(base_size = 12))
```


Indicator for ecological condition may be scaled against a reference value to become normalised. Actually, in SEEA EA, unscaled or raw indikcatr are not called indicators at all, but simply referred to as variables. But scaling variables against reference conditions can be done in many ways. Here are some exampels, for inspiration.


```{r}
set.seed(123)
```


Create random raw data for two variables
```{r}
dat <- data.frame(variable1 = rnorm(50, 10, 4), 
                  variable2 = rnorm(50, 30, 3))

temp <- data.table::melt(setDT(dat))
ggplot(temp,
       aes(x = value,
           y = variable,
           fill = stat(x)))+
  geom_density_ridges_gradient(scale = 3, size = 0.3)+
  scale_fill_gradient(low = 'red', high = 'green')
```

Lets say that the reference condition for variable1 to equal 16, and for variable 2 its 35

```{r}
dat$referenceHigh1 <- 16
dat$referenceHigh2 <- 35
```



## Linear, with a natural zero

We can scale our variable linearly, saying that all values above 12 represents 'perfect' condition. We also assume that the value zero is the worst possible condition for this variable. To do this we divide by the reference value, and truncate values above it.

```{r}
dat$indicator1 <- dat$variable1/dat$referenceHigh1
dat$indicator1[dat$indicator1>1] <- 1

dat$indicator2 <- dat$variable2/dat$referenceHigh2
dat$indicator2[dat$indicator2>1] <- 1
```

```{r}
temp <- melt(setDT(dat),
             measure.vars = c("indicator1", "indicator2"))
```


```{r}
ggplot(dat)+
  #geom_smooth(size=1.5, colour = "lightblue")+
  geom_point(
    aes(x = variable1, y = indicator1, fill = indicator1),
             colour="black",
             size=10,
             alpha=.7,
             shape=21)+
  geom_vline(xintercept = 12,
             linetype = "dashed",
             size =2)+
  
  geom_point(aes(x = variable2, y = indicator2, fill = indicator2),
             colour="black",
             size=10,
             alpha=.7,
             shape=21)+
  geom_vline(xintercept = 35,
             linetype = "dashed",
             size =2)+
  ylab("Indicator values")+
  xlab("Variable values")+
  
  scale_fill_gradient("Indicator values", low = 'red', high = 'green')
```

The two indicator become more comparable, at least around the reference value. But perhaps a variable value around 25 for variabel2 is qctually quite bad. Here we have assumed that the value zero is the worst possible value. That may not always be the case

## Linear, with defined lower limit
We can also say that the worst possible condition for thess variables are not zero, but for example say that it is 5 for variable1 and 28 for variable2.
```{r}
dat$referenceLow1 <- 5
dat$referenceLow2 <- 28
```


```{r}
dat$indicator1_LowHigh <- (dat$variable1-dat$referenceLow1)/(dat$referenceHigh1 - dat$referenceLow1)
dat$indicator1_LowHigh[dat$indicator1_LowHigh <0] <- 0
dat$indicator1_LowHigh[dat$indicator1_LowHigh >1] <- 1

dat$indicator2_LowHigh <- (dat$variable2-dat$referenceLow2)/(dat$referenceHigh2 - dat$referenceLow2)
dat$indicator2_LowHigh[dat$indicator2_LowHigh <0] <- 0
dat$indicator2_LowHigh[dat$indicator2_LowHigh >1] <- 1
```

```{r}
ggplot(dat)+
  geom_point(
    aes(x = variable1, y = indicator1_LowHigh, fill = indicator1_LowHigh),
             colour="black",
             size=10,
             alpha=.7,
             shape=21)+
  geom_vline(xintercept = 16,
             linetype = "dashed",
             size =2)+
  geom_vline(xintercept = 5,
             linetype = "dashed",
             size =2)+
  
  geom_point(aes(x = variable2, y = indicator2_LowHigh, fill = indicator2_LowHigh),
             colour="black",
             size=10,
             alpha=.7,
             shape=21)+
  geom_vline(xintercept = 35,
             linetype = "dashed",
             size =2)+
  geom_vline(xintercept = 28,
             linetype = "dashed",
             size =2)+
  ylab("Indicator values")+
  xlab("Variable values")+
  
  scale_fill_gradient("Indicator values", low = 'red', high = 'green')
```



## Break point scaling
We can also define a threshold for good ecological condition. Lets for example say that for variable1, ecosystem function really declines when the variable shows values below 9.

```{r}
dat$thr1 <- 9
```

```{r}
dat$indicator1_breakPoint <- dat$indicator1_LowHigh
dat$indicator1_breakPoint[dat$variable1 < dat$thr1] <- (1.6*dat$indicator1_LowHigh[dat$variable1 < dat$thr1])

dat$indicator1_breakPoint[dat$variable1 > dat$thr1] <- (dat$variable1[dat$variable1 > dat$thr1]-
                                                          dat$thr1[dat$variable1 > dat$thr1])/
                                                        (dat$referenceHigh1[dat$variable1 > dat$thr1] - 
                                                           dat$thr1[dat$variable1 > dat$thr1])+0.6

```


```{r}
ggplot(dat)+
  geom_point(
    aes(x = variable1, y = indicator1_breakPoint, fill = indicator1_breakPoint),
             colour="black",
             size=10,
             alpha=.7,
             shape=21)+
  geom_vline(xintercept = 16,
             linetype = "dashed",
             size =2)+
  geom_vline(xintercept = 5,
             linetype = "dashed",
             size =2)+
  geom_segment(aes(x = thr1, y = 0, xend = thr1, yend = 0.6),
               linetype="dashed",
               size=1.5)+
  geom_segment(aes(x = 0, y = 0.6, xend = thr1, yend = 0.6),
               linetype="dashed",
               size=1.5)+
  
  ylab("Indicator values")+
  xlab("Variable values")+
  
  scale_fill_gradient("Indicator values", low = 'red', high = 'green')
```

```{r}

```


